### 2015-06-08 chapters/ipa_meets_unicode.tex
starting to edit chapter 3

The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA
uses Latin-based glyphs like <a>, <b> or <p>. From this perspective, the IPA
seems to be just another orthographic tradition using Latin characters, all of
which do not get a special treatment within the Unicode Standard (just like
e.g.~the French, German, or Danish orthographic traditions do not have a special
status). On the other hand, the IPA uses many special symbols (like turned,
mirrored and/or extended Latin glyphs) not found in any other Latin-based
writing system. For this reason, and already in the first version of the Unicode
Standard (Version 1.0 from 1991), a special block with code points, called
\textsc{IPA Extensions} was included.

@@

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical, depending on the
font rendering). For example, a speaker of Russian should ideally not use the
<а> \textsc{cyrillic small letter a} at code point \uni{0430} for IPA
transcriptions, but instead use the <a> \textsc{latin small letter a} at code point
\uni{0061}, although visually they are mostly indistinguishable, and the
Cyrillic character is more easily typed on a Cyrillic keyboard.

Furthermore, even linguists are unlikely to distinguish between
the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
\textsc{latin small letter turned e} at \uni{01DD}. Conversely,
non-linguists are unlikely to distinguish any semantic difference between an
open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter
a} at \uni{0061}. But even among linguists this distinction leads to problems.
For example, as pointed out by \citet{Mielke2009}, there is a problem stemming
from the fact that about 75\% of languages are reported to have a five-vowel
system \citep{Maddieson1984}. Historically, linguistic descriptions tend not to
include precise audio recording and measurements of formants, so this may lead
one to ask if the many characters that are used in phonological description
reflects a transcriptional bias. The common use of <a> in transcriptions
could be in part due to the ease of typing the letter on an English keyboard (or
for older descriptions, the typewriter). We found it to be exceedingly rare that
a linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
/a/ as [+back] results from its behavior in certain phonological contexts.
For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
[+back] vowels), as opposed to /i/ and /e/.''} They simply use <a> as
long as there is no opposition to <ɑ>.\footnote{See Thomason's Language Log
post, ``Why I don't love the International Phonetic Alphabet'' at:
\url{http://itre.cis.upenn.edu/~myl/languagelog/archives/005287.html}.}

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that in most well-designed fonts the italics of <a> and <ɑ>
use the same glyph. If this distinction has to be kept upright in italics, the
only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ>. However, note that when the text is transferred to another font
(i.e.~one that includes real italics) and/or to another software application
(like Apple Pages, which does not perform slanting), then this visual appearance
will be lost. In this case we are thus still in the pre-Unicode situation in
which the choice of font and rendering software actually matters. The ideal
solution from a linguistic point of view would be the introduction of a new IPA
code point for a different kind of which explicitly specifies that it should
still be rendered as a double-story character when italicized. After informal
discussion with various Unicode players, our impression is that this highly
restricted problem is not sufficiently urgent to introduce even more <a>-like
characters in Unicode (which already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}

>>>>

The ambivalent nature of IPA glyphs arises because, on the one hand, the IPA
uses Latin-based glyphs like <a>, <b> or <p>. From this perspective, the IPA
seems to be just another orthographic tradition using Latin characters, all of
which do not get a special treatment within the Unicode Standard (just like
e.g.~the French, German, or Danish orthographic traditions do not have a special
status). On the other hand, the IPA uses many special symbols (like turned <ɐ>,
mirrored <ɘ> and/or extended <ɧ> Latin glyphs) not found in any other Latin-based
writing system. For this reason a special block with code points, called
\textsc{IPA Extensions} was included already in the first version of the Unicode
Standard (Version 1.0 from 1991).

@@

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical glyphs, depending
on the font rendering). For example, a user of a Cyrillic computer keyboard
should ideally not use the <а> \textsc{cyrillic small letter a} at code point
\uni{0430} for IPA transcriptions, but instead use the <a> \textsc{latin small
letter a} at code point \uni{0061}, although visually they are mostly
indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic
keyboard.

Furthermore, even linguists are unlikely to distinguish between
the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
\textsc{latin small letter turned e} at \uni{01DD}.

Conversely,
non-linguists are unlikely to distinguish any semantic difference between an
open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter
a} at \uni{0061}. But even among linguists this distinction leads to problems.
For example, as pointed out by \citet{Mielke2009}, there is a problem stemming
from the fact that about 75\% of languages are reported to have a five-vowel
system \citep{Maddieson1984}. Historically, linguistic descriptions tend not to
include precise audio recording and measurements of formants, so this may lead
one to ask if the many <a> characters that are used in phonological description
reflects a transcriptional bias. The common use of <a> in transcriptions
could be in part due to the ease of typing the letter on an English keyboard (or
for older descriptions, the typewriter). We found it to be exceedingly rare that
a linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
/a/ as [+back] results from its behavior in certain phonological contexts.
For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
[+back] vowels), as opposed to /i/ and /e/.''} They simply use <a> as
long as there is no opposition to <ɑ>.\footnote{See Thomason's Language Log
post, ``Why I don't love the International Phonetic Alphabet'' at:
\url{http://itre.cis.upenn.edu/~myl/languagelog/archives/005287.html}.}

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that in most well-designed fonts the italics of <a> and <ɑ>
use the same glyph. If this distinction has to be kept upright in italics, the
only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ> in italics. However, note that when the text is transferred to another
font (i.e.~one that includes real italics) and/or to another software
application (like Apple Pages, which does not perform slanting), then this
visual appearance will be lost. In this case we are thus still in the
pre-Unicode situation in which the choice of font and rendering software
actually matters. The ideal solution from a linguistic point of view would be
the introduction of a new IPA code point for a different kind of <a>, which
explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a> homographs in Unicode (which
already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}
